Meta-analytic approaches and 'new' effect sizes to account for heterogeneous effects in comparative physiology

### 2021-06-30 MS/ms.Rmd
MOre fixes and dealing with a few more comments from Shin

Of course, there are certain types of heterogeneity that transcend different questions. In other words, if we compare different types of 'apples' is it still fair to synthesize these studies to understand overall effects? This situation might be analogous to a meta-analysis in comparative physiology that attempts to understand the effects of temperature on fish growth. <!--- taxnomically speaking - comparing different applies are comparing different populations of the same species of fish so not quite sure whether this is quite correct - I am being of course a bit pedantic---> Studies might rear populations of fish under different temperatures, but in the end, they are still all fish, governed by the same thermodynamic effects on growth. In this situation, we might still be interested in understanding overall effects, but we may question whether such a conclusion is sensible given the diversity of temperatures applied. Such a study does not preclude us from also exploring important moderators or predictors of variability in effects across studies.....

@@

As discussed above, it is clear that different types of questions and experimental approaches will be more or less susceptible to the 'apples and oranges' problem. <!--- not sure whether it is a good idea to highlight it as a problem as said earlier - explaining hetero is what's interesting and even the main purpose of meta-analysis for comparative physiologists - they are not the problem but they are what we are going to investigate. I do not think 'susceptible' is a good word here given the average hetero = > 90% - you are nearly 100% guaranteed to have hetero and this is what's interesting ---> As such, before embarking on our path to 'meta-analytic enlightenment', we conducted a literature survey to understand how widespread the problem may be in comparative physiology and gauge the ways in which it is currently being addressed, if at all. To accomplish this goal, we searched for meta-analyses and comparative analyses that tackled questions using physiological data compiled from animal studies across multiple species. We were specifically interested in asking: 1) what type of effect size was used within the quantitative synthesis? 2) based on the methods provided, was the effect size likely susceptible to continuous variation in how treatments were applied (e.g., for contarst-based effects) or the conditions under which traits were measured (i.e., for correlatioons) across studies? 3) If so, was the variation resulting from treatment differences accounted for in the effect size or during analysis?, and finally, 4) how were effect sizes analysed in the quantitative synthesis, were these weighted analyses, and what program was used for analysis? We also recorded whether publication bias and sensitivity analyses were conducted, if data were publicly available, where it was stored, and if so, if it was usable (i.e., in a machine readable format).

@@

All title, abstract and full text screening was done by two people (DWAN and ML). Any conflicts or disagreements were resolved by discussion and a consensus on a paper was reached. Out of the 417 papers, 79 full texts were screened and we included a total of 62 for final data extraction. For full details on the search strings used, dates searched, the number of papers found, and our PRISMA flowchart (along with reasons for exclusion at the full text stage) we refer readers to the *Supplementary Materials*. To address our questions, we categorised effect sizes as falling into the following categories: Response Ratio (lnRR), Standardised Mean Differences (SMD; e.g., Hedges d, g or Cohen's d), Correlation or Fisher's Z -transformed Correlation, Regression Slopes (Slopes), Risk Ratios, or whether the paper analysed the raw data or summary statistics directly (i.e., means/proportions) (which we label as raw/mean/proportion). For all other effects we labelled as 'other'. To determine whether effect sizes were likely to vary based on some underlying continuous variable or were dependent on the context in which traits were measured, we assessed whether authors: 1) acknowledged this explicitly in the paper or 2) accounted for it in their models. If it was not clear, two of us (DWAN & PP) discussed the paper, and made a decision. For example, trait means, such as ectothermic metabolic rate (i.e., raw/mean/proportion-based effect size), will vary depending on the temperature being measured, the magnitude of differences between two treatments (e.g, contrast-based effect sizes - SMD or lnRR) might depend on the temperature or dosage used in treatments, and the strength of relationship between two variables (e.g., correlation between behaviour and metabolism) might depend on the temperature context under which metabolic rate or behaviour were measured. Given the high heterogeneity in effects used in comparative physiology, we expected this to be common to most effect sizes, however, it has not been formally quantified.

@@

Prior to showing how the relevant $Q_{10}$ effect size can be calculated it can be helpful to understand its similarities to *lnRR*. The *lnRR* described by @Hedges1999 and extended by @Lajeunesse2011 and can be calculated as follows <!---put see also Senior et al. 2020 RSM as well - there we expand the 2nd order delta method stuff - so unbiased formlas for all lnRR, lnVR etc is a bit different - we do not want to put it here but mention by the way - Lajeuness 2011 is not the one I would cite Lajeuness 2015 is more relevant ---> :

>>>>

Of course, there are certain types of heterogeneity that transcend different questions. In other words, if we compare different types of 'apples' is it still fair to synthesize these studies to understand overall effects? This situation might be analogous to a meta-analysis in comparative physiology that attempts to understand the effects of temperature on fish growth across different populations of a widely distributed (cosmopoliton) species. Studies might rear populations of fish under different temperatures, but in the end, they are still all the same fish species, governed by the same thermodynamic effects on growth. In this situation, we might still be interested in understanding overall effects, but we may question whether such a conclusion is sensible given the diversity of temperatures applied. Such a study does not preclude us from also exploring important moderators or predictors of variability in effects across studies.....

@@

As discussed above, it is clear that different types of questions and experimental approaches will mix 'apples and oranges' to varying extents, and indeed, this heterogenity (or basket of fruit) is often the main interest to comparative biologists. To better understand the different types of effect sizes being used, thier susceptability to known sources of methodological / environmental variablity across studies, and how these sources are controlled we conducted a literature survey. To accomplish this goal, we searched for meta-analyses and comparative analyses that tackled questions using physiological data compiled from animal studies across multiple species. We were specifically interested in asking: 1) what type of effect size was used within the quantitative synthesis? 2) based on the methods provided, was the effect size likely susceptible to continuous variation in how treatments were applied (e.g., for contarst-based effects) or the conditions under which traits were measured (i.e., for correlatioons) across studies? 3) If so, was the variation resulting from treatment differences accounted for in the effect size or during analysis?, and finally, 4) how were effect sizes analysed in the quantitative synthesis, were these weighted analyses, and what program was used for analysis? We also recorded whether publication bias and sensitivity analyses were conducted, if data were publicly available, where it was stored, and if so, if it was usable (i.e., in a machine readable format).

@@

All title, abstract and full text screening was done by two people (DWAN and ML). Any conflicts or disagreements were resolved by discussion and a consensus on a paper was reached. Out of the 417 papers, 79 full texts were screened and we included a total of 62 for final data extraction. For full details on the search strings used, dates searched, the number of papers found, and our PRISMA flowchart (along with reasons for exclusion at the full text stage) we refer readers to the *Supplementary Materials*. To address our questions, we categorised effect sizes as falling into the following categories: Response Ratio (lnRR), Standardised Mean Differences (SMD; e.g., Hedges d, g or Cohen's d), Correlation or Fisher's Z -transformed Correlation, Regression Slopes (Slopes), Risk Ratios, or whether the paper analysed the raw data or summary statistics directly (i.e., means/proportions) (which we label as raw/mean/proportion). For all other effects we labelled as 'other'. To determine whether effect sizes were likely to vary based on some underlying continuous variable or were dependent on the context in which traits were measured, we assessed whether authors: 1) acknowledged this explicitly in the paper or 2) accounted for it in their models. If it was not clear, two of us (DWAN & PP) discussed the paper, and made a decision. For example, trait means, such as ectothermic metabolic rate (i.e., raw/mean/proportion-based effect size), will vary depending on the temperature being measured, the magnitude of differences between two treatments (e.g, contrast-based effect sizes - SMD or lnRR) might depend on the temperature or dosage used in treatments, and the strength of relationship between two variables (e.g., correlation between behaviour and metabolism) might depend on the temperature context under which metabolic rate or behaviour were measured. Given the high heterogeneity in effects used in comparative physiology, we expected this to be common to most effect sizes, however, it has not been formally quantified. We categorised the types of staitsical models fit in each study as: 1) fixed-effect meta-analysis; 2) random effects meta-analysis (i.e., intercept only model, but has random study effects); 3) multi-level meta-analysis (i.e., intercept only model, but has multiple random effect levels such as study, species, observation-level random effects); 4) meta-regression models (i.e., a multilevel or random effect model with fixed effects - including random effects); 5) weighted regression models, inluding mixed effect models; and 6) linear regression models (including linear and generalised linear mixed effect models, and phylogenetic least squares models). Notably, only model types 1-5 are weighted models.

@@

Prior to showing how the relevant $Q_{10}$ effect size can be calculated it can be helpful to understand its similarities to *lnRR*. The *lnRR* described by @Hedges1999 and extended by @Lajeunesse2015 and can be calculated as follows [but see also @Senior2020]:
